Practical Optional Types for Clojure

### 2015-02-27 formal-model.tex
move T-If to prose

%TODO EXPLAIN what is a proposition?
The first insight about occurrence typing is that
logical formulas
can be used to represent type information about our programs
by relating parts of the runtime environment to types.
Occurrence typing uses propositional logic 
\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---
in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard logical connectives: implications, conjunctions,
disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a sequence of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we use the proposition {\isprop{\NumberFull}{\x{}}}.

Given a set of propositions, we can use logical reasoning to derive
new information about our programs
with the judgement \inpropenv{\propenv{}}{\prop{}}.
In addition to the standard rules for the logical connectives, the key
rule is L-Update, which combines multiple propositions about the same variable.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

@@

Figure~\ref{main:figure:typingrules} contains the core typing rules.
The key rule for reasoning about conditional control flow is
T-If. The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
are used as assumptions in the then and else branch respectively.
If the result of the \ifliteral{} is true, then the value either
came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
which implies \thenprop{\prop{3}} is true. 
The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
similarly.
The T-Local rule connects the type system to the proof system over type propositions
to derive a type for a variable.
Using this rule, the type system can then appeal to L-Update to refine the type
assigned to \x{}.

@@

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    {\TLocal}

    {\TTrue}

    {\TFalse}

    {\TNil}

    %{\TDo}
    %{\TClass}
    {\TIf}
    {\TLet}
    \begin{array}{c}
      {\TConst}\\\\
      {\TKw}\\\\
      {\TClass}
    \end{array}

    {\TApp}

    {\TAbs}

    %{\TError}

    {\TSubsume}
  \end{mathpar}
  \caption{Typing rules}
  \label{main:figure:typingrules}
\end{figure*}

>>>>

%TODO EXPLAIN what is a proposition?
The first insight about occurrence typing is that
logical formulas
can be used to represent type information about our programs
by relating parts of the runtime environment to types
via propositional logic.
\emph{Type Propositions} \prop{} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard logical connectives: implications, conjunctions,
disjunctions, and the trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a set of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we use the proposition {\isprop{\NumberFull}{\x{}}}.

Given a set of propositions, we can use logical reasoning to derive
new information about our programs
with the judgement \inpropenv{\propenv{}}{\prop{}}.
In addition to the standard rules for the logical connectives, the key
rule is L-Update, which combines multiple propositions about the same variable,
allowing us to refine its type.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\Nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

@@

Figure~\ref{main:figure:typingrules} contains the core typing rules.
The key rule for reasoning about conditional control flow is
T-If.

\begin{mathpar}
\TIf{}
\end{mathpar}

The propositions of the test \thenprop{\prop{1}} and \elseprop{\prop{1}}
are used as assumptions in the then and else branch respectively.
If the result of the \ifliteral{} is true, then the value either
came from \e{2}, in which case \thenprop{\prop{2}} is true, or from \e{3},
which implies \thenprop{\prop{3}} is true. 
The else proposition is \orprop{\elseprop{\prop{2}}}{\elseprop{\prop{3}}} 
similarly.
The T-Local rule connects the type system to the proof system over type propositions
to derive a type for a variable.
Using this rule, the type system can then appeal to L-Update to refine the type
assigned to \x{}.

@@

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    %{\TDo}
    %{\TClass}
    %{\TIf}
    {\TAbs}
    {\TSubsume}
    \begin{array}{c}
      {\TConst}\\\\
      {\TKw}\\\\
      {\TClass}
    \end{array}
    \begin{array}{c}
      {\TTrue}\\\\
      {\TFalse}\\\\
      {\TNil}
    \end{array}

    {\TLet}
    {\TLocal}

    {\TApp}
    %{\TError}

  \end{mathpar}
  \caption{Typing rules}
  \label{main:figure:typingrules}
\end{figure*}

### 2015-02-27 intro.tex
move T-If to prose

The \clj{parent} function has the type 
$$
\clj{['{:file (U nil File)} -> (U nil Str)]}
$$
which means that it takes a hash table whose \clj{:file} key maps to either
\clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
\clj{String}. The \clj{parent} function uses the \clj{:file} keyword
as an accessor to get the file, checks that it isn't \clj{nil}, and
then obtains the parent by making a Java method call.
%
The annotation \clj{^File} is a type hint, which instructs the Clojure
compiler (running prior to Typed Clojure typechecking) to statically
resolve method calls to the relevant object using the Java \clj{File}
class, rather than using reflective method calls at runtime.

>>>>

The \clj{parent} function has the type 
$$
\clj{['{:file (U nil File)} -> (U nil Str)]}
$$
which means that it takes a hash table whose \clj{:file} key maps to either
\clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
\clj{String}. The \clj{parent} function uses the \clj{:file} keyword
as an accessor to get the file, checks that it isn't \clj{nil}, and
then obtains the parent by making a Java method call.
%
The annotation \clj{^File f} is a type hint on \clj{f}, which instructs the Clojure
compiler (running prior to Typed Clojure typechecking) to statically
resolve the \clj{getParent} 
call to \clj{File}'s \clj{getParent} method with signature \java{String getParent();}, rather than using reflective method calls at runtime.

### 2015-02-27 overview.tex
move T-If to prose

Here \clj{defalias} defines a type abbreviation called \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
\clj{an-exp} is a function verified to return an \clj{Expr}.

@@

\begin{exmp}
\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
\end{exmp}

@@

If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
we remove
the \clj{:do} and \clj{:const}
Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
and we are left with an \clj{:if} Expr.
On line 8,
we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
Line 9 we know \clj{m}
we remove \clj{:const} because it contradicts \clj{(= (:op m) :do)} being true,
and we know \clj{m} is a \clj{:do} Expr.
Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
so we are left with \clj{m} being a \clj{:const} Expr.

@@

The Clojure compiler uses the type hint
to statically resolve 
the method call to the
\mintinline{java}{public String getParent()}
method of \clj{java.io.File}.
The method is unambiguous but type checking fails---Typed Clojure
considers \clj{f} to be of type \clj{Any}, which is unsafe 
to use as the target of even a resolved method.
We annotate the function parameter with type \clj{File}.

@@

The type hinting system and Typed Clojure's static type checking
are separate, the latter predating the former by several years.
The interaction between them is often not as obvious, for example type hints
do not need to take \clj{nil} into account, while Typed Clojure guarantees
null-pointer exceptions are impossible.

@@

To guarantee \java{null} is never accidentally leaked into a Typed Clojure program,
we must assume methods are nullable.

@@

In contrast, JVM invariants guarantee that  constructor cannot return \java{null},
so we are safe to assume constructors are non-nullable.

@@

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispath table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path (File. "dir/a"));=> IllegalArgumentException
\end{minted}

@@

\begin{figure}
\begin{exmp}
\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
\label{example:incmap}
\end{exmp}
\end{figure}

>>>>

Here \clj{defalias} defines a type abbreviation called \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
A quoted keyword in a type such as \clj{':if} is a singleton type that contains just the keyword.
\clj{an-exp} is a function verified to return an \clj{Expr}.

@@

\begin{exmp}
\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
\label{example:circleci}
\end{exmp}

@@

If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
we remove
the \clj{:do} and \clj{:const}
Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
and we are left with an \clj{:if} Expr.
On line 8,
we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
Line 9 know we can
remove the \clj{:const} Expr from the type of \clj{m} because it contradicts \clj{(= (:op m) :do)} being true,
and we know \clj{m} is a \clj{:do} Expr.
Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
so we are left with \clj{m} being a \clj{:const} Expr.

@@

The Clojure compiler uses the type hint
to statically resolve 
the method call to the
\mintinline{java}{public String getParent()}
method of \clj{java.io.File}.
The call to \clj{getParent} is unambiguous at runtime but type 
checking fails---Typed Clojure
considers \clj{f} to be of type \clj{Any}, which is unsafe 
to use as the target of even a resolved method.
If instead we annotate the function parameter with type \clj{File},
we get a static type error.

@@

% type hints give no guarantees
% TODO

The type hinting system and Typed Clojure's static type checking
are separate, the latter predating the former by several years.
The interaction between them is often not as obvious, for example Typed Clojure
has an explicit type for \clj{null}
null-pointer exceptions are impossible.

do not need to take \clj{nil} into account,

@@

Reference types in Java are nullable, so to guarantee a method call does not
leak \java{null} into a Typed Clojure program we
must assume methods can return \clj{nil}.

@@

In contrast, JVM invariants guarantee that  constructors cannot return \java{null},
so we are safe to assume constructors are non-nullable.

@@

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path (File. "dir/a"));=> IllegalArgumentException
\end{minted}

@@

\begin{exmp}
\inputminted[firstline=5,lastline=18]{clojure}{code/demo/src/demo/eg5.clj}
\label{example:incmap}
\end{exmp}
